---
title: "Exploratory analysis for local findings"
author: "Dillon Bergin, MuckRock Data Reporter"
date: "09/21/2022"
output: 
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
---

```{r loading, warning=F, message=F}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(lubridate)
library(rio)
library(DT)
library(tidycensus)
library(tigris)
library(tmap)
```

```{r results=F, warning=F, message=F}
unweighted_grids <- read_csv(here("data", "processed", "output_etl", "ca_county_grid.csv"))
weighted_grids <-read_csv(here("data", "processed", "childs_county_level_results", "annual_county_smokePM.csv"))
```

```{r results=F, warning=F, message=F}
ca_fips <- counties("CA") %>% 
  select(county = NAME, county_fips = GEOID) %>% 
  st_drop_geometry()

ca_county_geoms <- counties("CA") %>% 
  select(county = NAME) 
```

```{r results=F, warning=F, message=F}
weighted_grids_join <- 
  weighted_grids %>% 
  inner_join(ca_fips, by = "county_fips") %>% 
  select(county, year, pm25 = annual_avg_smokePM)

```
# Across all years and months, which counties have most wildfire pm? 
### Take-away
- Won't surprise anyone, looks like maps we've seen before

```{r warning=F, message=F}

average_pm <- weighted_grids_join %>% 
  group_by(county) %>% 
  summarize(average_pm = mean(pm25))


county_sf <- average_pm %>% 
inner_join(ca_county_geoms, by = "county") %>% 
  st_sf()

tm_shape(county_sf) +
  tm_polygons("average_pm",
              style = "quantile",
              title = "Average Wildfire PM Across All Years") + 
              tm_text("county", size = 1/4)
```

# Do these counties also have the most days with smoke overhead? Is there a difference between the two? 
### Take-away
- Big differences look like Fresno and Tulare, as well as Imperial on the bottom. Is this downwind smoke? 

```{r warning=F, message=F}
smoke_days_by_year <- unweighted_grids %>% 
  group_by(county, year = floor_date(date, "year")) %>% 
  summarize(days = n_distinct(date)) %>% 
  mutate(year = as.numeric(format(year,'%Y'))) %>% 
  select(county, year, days) 

smoke_days_average <- smoke_days_by_year %>% 
  group_by(county) %>% 
  summarize(average_smoke_days = round(mean(days), digits = 2))

smoke_days_sf <- smoke_days_average %>% 
inner_join(ca_county_geoms, by = "county") %>% 
  st_sf()

tm_shape(smoke_days_sf) +
  tm_polygons("average_smoke_days",
              style = "quantile",
              title = "Average Days of Smoke Plumes Overhead") + 
              tm_text("county", size = 1/4)


```



### Data to explore average smoke days (overhead) across all years 
```{r warning=F, message=F}

datatable(smoke_days_average, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',

                            options = list(
                                paging = TRUE,
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))
  

```

### Data to explore average smoke days (overhead) in yearly changes 
```{r warning=F, message=F}
smoke_days_by_year <- unweighted_grids %>% 
  group_by(county, year = floor_date(date, "year")) %>% 
  summarize(days = n_distinct(date)) %>% 
  mutate(year = as.numeric(format(year,'%Y'))) %>% 
  select(county, year, days) 

smoke_days_by_year_wide <- smoke_days_by_year %>% 
  pivot_wider(names_from = year, values_from = days)

datatable(smoke_days_by_year_wide, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',

                            options = list(
                                paging = TRUE,
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))
```


# Which months were the worst? 
### Take-away
- Caveat: this is unweighted, so we won't be making the claim which days were the "worst for people's health"
- Top five worst months? August, September, October of 2020 along with August and November of 2018

```{r warning=F, message=F}
worst_months <- unweighted_grids %>% 
  mutate(date = floor_date(as_date(date), "month")) %>% 
  group_by(date) %>% 
  summarize(avg_wildfire_pm = round(mean(pm25), digits = 2)) %>% 
  mutate(date = format(date, "%B %Y"))

datatable(worst_months, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',

                            options = list(
                                paging = TRUE,
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))
  
  
```

# Which days were the worst and where? 
### Take-away
- this seems interesting, especially those three very intense days in Mono County. Estimates reach over 300 micrograms for three days in a row. I'm sure Mono is less populated than other areas of the state, but honing on days like these could be a way to turn back to health experts and say - how bad is breathing in that much PM over multiple days? How dangerous is for people with asthma or other respiratory diseas? 


```{r warning=F, message=F}
worst_days <- unweighted_grids %>% 
  group_by(county, date) %>% 
  summarize(avg_wildfire_pm = round(mean(pm25), digits = 2)) %>% 
  arrange(desc(avg_wildfire_pm)) %>%
  slice(1:100) 

datatable(worst_days, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',

                            options = list(
                                paging = TRUE,
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))

```

# How did fall 2020 look by county? 

```{r warning=F, message=F}

worst_months_by_county <- unweighted_grids %>% 
  mutate(date = floor_date(as_date(date), "month")) %>% 
  group_by(date, county) %>% 
  summarize(avg_wildfire_pm = round(mean(pm25), digits = 2)) %>% 
  mutate(date = format(date, "%B %Y")) 

fall_2020_sf <- worst_months_by_county %>% 
  filter(date %in% c("August 2020", "September 2020", "October 2020")) %>% 
  inner_join(ca_county_geoms, by = "county") %>% 
  st_sf()

tm_shape(fall_2020_sf) +
  tm_polygons("avg_wildfire_pm",
              style = "quantile",
              title = "Total Increase in Smoke PM") + 
              tm_text("county", size = 1/4) +
              tm_facets("date")


```



