---
title: "Final analysis for local findings"
author: "Dillon Bergin, MuckRock Data Reporter"
date: "09/23/2022"
output: 
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
--- 

```{r loading, warning=F, message=F}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(lubridate)
library(rio)
library(tidycensus)
library(tigris)
library(tmap)
library(zoo)
library(DT)
```

```{r results=F, warning=F, message=F}
unweighted_grids <- read_csv(here("data", "processed", "output_etl", "ca_county_grid.csv"))
weighted_grids <- read_csv(here("data", "processed", "childs_county_level_results", "annual_county_smokePM.csv"))
#grid_pop <- read_csv(here("data", "raw", "childs_et_al_study_data", "grid_id_pop.csv"))
extreme_day_pop_estimates <- read_csv(here("data", "processed", "childs_county_level_results", "ca_extreme_days_pop_estimates.csv"))
```


FINDING #1: 21-fold increase

“That includes 21 times more Californians than a decade ago, scattered among vulnerable communities from the Oregon border, to the foothills of the Sierra Nevadas and down through the Central Valley.”

```{r}
  
# Double check there truly isn't a day over 100 PM2.5 in all of 2006, 2010, or 2011
#years_2006_2010 <- unweighted_pop %>% 
  #(year = floor_date(as_date(date), "year")) %>% 
  #mutate(year = format(date, "%Y")) %>% 
  #filter(year %in% c(2006, 2010))

change <- extreme_day_pop_estimates %>% 
  select(1,3) %>% 
  pivot_wider(names_from = year, values_from = pop_over100) %>% 
  mutate(total_then = sum(`2006`,`2007`, `2008`, `2009`, `2010`)) %>%  
  mutate(total_now = sum(`2016`, `2017`, `2018`, `2019`, `2020`)) %>% 
  select("total_now", "total_then") %>% 
  mutate(fold_change = (total_now/total_then))

extreme_2015_2020 <- extreme_day_pop_estimates %>% 
  slice(10:15)

median_2006_2020 <- median(extreme_day_pop_estimates$pop_over100)
median_2015_2020 <- median(extreme_2015_2020$pop_over100)




```

Populations exposed to at least one day of smoke over 100 micrograms (data supplied by Childs)
```{r}

datatable(extreme_day_pop_estimates, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',

                            options = list(
                                paging = TRUE,
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))

```

Fold change 
```{r}
datatable(change, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',

                            options = list(
                                paging = TRUE,
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))

```


# FINDING 2: Smoke days in 2020 and average PM 

“Childs and her colleagues found that in 2020, when five of the largest fires in history raced across the state, smoke plumes hung overhead for more days than any year in the last decade. During that time, they estimate that ground monitors captured wildfire-specific pollution three times higher than normal."

Smoke plume days 
```{r}

# Smoke plume days 
annual_av_days <-  unweighted_grids %>% 
  mutate(year = floor_date(as_date(date), "year")) %>% 
  mutate(year = format(date, "%Y")) %>% 
  group_by(year) %>% 
  summarize(smoke_days = n_distinct(date))
  
annual_day_comparison <- annual_av_days %>% 
  pivot_wider(names_from = year, values_from = smoke_days) %>% 
  mutate(avg_decade = sum(`2010`, `2011`, `2012`, `2013`, `2014`, `2015`, `2016`, `2017`, `2018`, `2019`)/10) %>% 
  select(`2020`, avg_decade, `2018`, `2008`)

datatable(annual_day_comparison, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',

                            options = list(
                                paging = TRUE,
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))


```


Average PM was three times higher than normal 
```{r}
# Average annual unweighted PM

unweighted_year_avg <- unweighted_grids %>% 
  mutate(year = floor_date(as_date(date), "year")) %>% 
  mutate(year = format(date, "%Y")) %>% 
  group_by(year) %>% 
  summarize(avg = mean(pm25))

comparison_unweighted_year <- unweighted_year_avg %>% 
  pivot_wider(names_from = year, values_from = avg) %>% 
  mutate(avg_decade = sum(`2010`, `2011`, `2012`, `2013`, `2014`, `2015`, `2016`, `2017`, `2018`, `2019`)/10) %>% 
  select(`2020`, avg_decade, `2018`, `2008`)


datatable(comparison_unweighted_year, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',

                            options = list(
                                paging = TRUE,
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))


```



# Finding three: decadal change 
"In the decade before 2015, an average of 35,000 Californians a year lived in areas where they were exposed to dangerous smoke. By the end of 2020, that number had doubled."



```{r}
  
# Double check there truly isn't a day over 100 PM2.5 in all of 2006, 2010, or 2011
#years_2006_2010 <- unweighted_pop %>% 
  #(year = floor_date(as_date(date), "year")) %>% 
  #mutate(year = format(date, "%Y")) %>% 
  #filter(year %in% c(2006, 2010))

decades_change <- extreme_day_pop_estimates %>% 
  select(1,3) %>% 
  pivot_wider(names_from = year, values_from = pop_over100) %>% 
  mutate(total_then = sum(`2006`,`2007`, `2008`, `2009`, `2010`)) %>%  
  mutate(total_now = sum(`2016`, `2017`, `2018`, `2019`, `2020`)) %>% 
  select("total_now", "total_then") %>% 
  mutate(fold_change = (total_now/total_then))

extreme_2015_2020 <- extreme_day_pop_estimates %>% 
  slice(10:15)

median_2006_2020 <- median(extreme_day_pop_estimates$pop_over100)
median_2015_2020 <- median(extreme_2015_2020$pop_over100)




```

Populations exposed to at least one day of smoke over 100 micrograms (data supplied by Childs)
```{r}

datatable(extreme_day_pop_estimates, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',

                            options = list(
                                paging = TRUE,
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))

```

Change for the given time period 
```{r}
datatable(change, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',

                            options = list(
                                paging = TRUE,
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))

```




# Finding 4: Northern California counties  


```{r results=F, warning=F, message=F}
weighted_grids_join <- 
  weighted_grids %>% 
  inner_join(ca_fips, by = "county_fips") %>% 
  select(county, year, pm25 = annual_avg_smokePM)

ca_county_geoms <- counties("CA") %>% 
  select(county = NAME) 

weighted_all_years_sf <- weighted_grids_join %>% 
  group_by(county, year) %>% 
  summarize(annual_avg = mean(pm25))

weighted_all_years_sf <- weighted_grids_join %>% 
  group_by(county, year) %>% 
  summarize(annual_avg = mean(pm25)) %>% 
  inner_join(ca_county_geoms, by = "county") %>% 
  st_sf()

tm_shape(weighted_all_years_sf) +
  tm_polygons("annual_avg",
              title = "decile") + 
              tm_text("county", size = 1/4)

```




FINDING #5: Pushing EPA standards 
```{r}

# Average PM2.5 per year across whole state

year_avg_pm <- unweighted_grids %>% 
  mutate(year = floor_date(as_date(date), "year")) %>% 
  mutate(year = format(date, "%Y")) %>% 
  group_by(year) %>% 
  summarize(avg = sum(pm25)/365)


comparison_year_avg_pm <- year_avg_pm %>% 
  pivot_wider(names_from = year, values_from = avg) %>% 
  mutate(now_avg = sum(`2016`, `2017`, `2018`, `2019`, `2020`)/5) %>% 
  mutate(then_avg = sum(`2006`, `2007`, `2008`, `2009`, `2010`)/5) %>% 
  select(then_avg, now_avg)

# Average PM2.5 per year by county 

# Get Fips codes from Tigris package 
ca_fips <- counties("CA") %>% 
  select(county = NAME, county_fips = GEOID) %>% 
  st_drop_geometry()

# Join with grid data 
weighted_grids_join <- 
  weighted_grids %>% 
  inner_join(ca_fips, by = "county_fips") %>% 
  select(county, year, pm25 = annual_avg_smokePM)

# Group weighted grids by county 
county_weighted <- weighted_grids_join %>% 
  group_by(county) %>% 
  summarize(average = mean(pm25))

nor_counties <-  county_unweighted %>% 
  filter(county %in% c("Trinity", "Siskiyou", "Shasta", "Tehama", "Glenn"))

nor_cal_mean <- mean(nor_counties$average)
all_mean <- mean(county_unweighted$average)
diff = (nor_cal_mean - all_mean)/all_mean

# Group grids by county unweighted 
county_unweighted <- unweighted_grids %>% 
  mutate(year = floor_date(as_date(date), "year")) %>% 
  mutate(year = format(date, "%Y")) %>% 
  group_by(year, county)
  summarize(average = sum(pm25)/365)

# Filter to San Joaquin unweighted 
san_joaquin_unweighted <-  county_unweighted %>% 
  filter(county %in% c("San Joaquin", "Stanislaus", "Merced", "Madera", "Fresno", "Kings", "Tulare", "Kern"))

mean_san_joaquin_unweighted <- mean(san_joaquin_unweighted$average)

# Filter to just San Joaquin weighted 
san_joaquin_weighted <-  county_weighted %>% 
  filter(county %in% c("San Joaquin", "Stanislaus", "Merced", "Madera", "Fresno", "Kings", "Tulare", "Kern"))

```



Visualization: Rolling Average of Population Exposure Change 

```{r}
rolling_av <- extreme_day_pop_estimates %>% 
  select(1,3) %>% 
  arrange(year) %>% 
  mutate(roll_mean = rollmean(pop_over100, k = 10, fill = NA, align = "right")) %>% 
  slice(10:15)


rolling_av %>% 
  ggplot(aes(x = year, y = roll_mean)) +
  geom_line()

```


Visualization: 


```{r}
weighted_counties_sf <- weighted_grids_join %>% 
  inner_join(ca_county_geoms, by = "county") %>% 
  st_sf()


small_multiples <- tm_shape(weighted_counties_sf) +
  tm_polygons("pm25",
              style = "quantile",
              title = "Impact of Wildfire Smoke on Californian's Air") + 
  tm_text("county", size = 1/4) +
  tm_facets("year")
```


```{r}

county_unweighted_daily <- unweighted_grids %>% 
  group_by(date, county) %>% 
  summarize(total_pm25 = mean(pm25))

county_unweighted_yearly <- county_unweighted_daily %>% 
  mutate(date = floor_date(date, "year")) %>% 
  mutate(year = format(date, "%Y")) %>% 
  select(year, county, total_pm25) %>% 
  group_by(year, county) %>% 
  summarize(average = mean(total_pm25))
    
  
county_unweighted_all_years <- county_unweighted_yearly %>% 
  group_by(county) %>% 
  summarize(average = mean(average))

san_joaquin_unweighted_2015_2020 <-  county_unweighted_yearly %>%
  filter(county %in% c("San Joaquin", "Stanislaus", "Merced", "Madera", "Fresno", "Kings", "Tulare", "Kern")) %>% 
  filter(year %in% c("2016", "2017", "2018", "2019", "2020")) %>% 
  group_by(county) %>% 
  summarize(average = mean(average))

mean <- mean(san_joaquin_unweighted_2015_2020$average) 

```




